---
title: "K-Means and K-Medoids Clustering on Simulated Data"
author: "Halil Bisgin"
output: html_document
---

## Introduction

In this analysis, we simulate data to apply both **K-Means** and **K-Medoids** clustering algorithms, compare their results, and visualize the clustering outcomes.

We also measure the time taken for each clustering method.

## Load Required Libraries

```{r setup, message=FALSE, warning=FALSE}
# Install libraries if not already installed
install.packages("cluster")
install.packages("factoextra")
install.packages("ggplot2")
install.packages("tictoc")

# Load necessary libraries
library(cluster)      # For K-Medoids (PAM)
library(factoextra)   # For visualization
library(ggplot2)      # For plotting
library(tictoc)       # For measuring execution time
```

```{r data prep}
# Set random seed for reproducibility
set.seed(42)

# Create 3 clusters with normal distribution
n <- 300  # Total number of data points
cluster1 <- data.frame(x = rnorm(n/3, mean = 2, sd = 0.5), y = rnorm(n/3, mean = 2, sd = 0.5))
cluster2 <- data.frame(x = rnorm(n/3, mean = 6, sd = 0.5), y = rnorm(n/3, mean = 6, sd = 0.5))
cluster3 <- data.frame(x = rnorm(n/3, mean = 10, sd = 0.5), y = rnorm(n/3, mean = 2, sd = 0.5))

# Combine clusters into one dataset
data <- rbind(cluster1, cluster2, cluster3)
colnames(data) <- c("feature1", "feature2")

# Normalize the data
scaled_data <- scale(data)

```
# Apply K-Means Clustering

```{r k-means}
tic("K-Means")  # Start timer
kmeans_result <- kmeans(scaled_data, centers = 3, nstart = 10)
toc()  # Stop timer
```

# Apply K-Medoids Clustering
```{r k-medoids}
tic("K-Medoids")  # Start timer
kmedoids_result <- pam(scaled_data, 3)
toc()  # Stop timer

```

# Prep data for visualization
```{r get memberships }
data$KMeans_Cluster <- as.factor(kmeans_result$cluster)
data$KMedoids_Cluster <- as.factor(kmedoids_result$clustering)
```

# Plotting k-means clusters
```{r vis kmeans}
p1 <- ggplot(data, aes(x = feature1, y = feature2, color = KMeans_Cluster)) +
  geom_point(size = 3) +
  ggtitle("K-Means Clustering") +
  theme_minimal()

# Print the K-Means plot
print(p1)
```

# Plotting k-medoids clusters
```{r kmedoids}
p2 <- ggplot(data, aes(x = feature1, y = feature2, color = KMedoids_Cluster)) +
  geom_point(size = 3) +
  ggtitle("K-Medoids Clustering") +
  theme_minimal()

# Print the K-Medoids plot
print(p2)
```
```{r}
# Function to generate data
generate_data <- function(n) {
  cluster1 <- data.frame(x = rnorm(n/3, mean = 2, sd = 0.5), y = rnorm(n/3, mean = 2, sd = 0.5))
  cluster2 <- data.frame(x = rnorm(n/3, mean = 6, sd = 0.5), y = rnorm(n/3, mean = 6, sd = 0.5))
  cluster3 <- data.frame(x = rnorm(n/3, mean = 10, sd = 0.5), y = rnorm(n/3, mean = 2, sd = 0.5))
  data <- rbind(cluster1, cluster2, cluster3)
  colnames(data) <- c("feature1", "feature2")
  return(scale(data))  # Normalize the data
}

# Define different dataset sizes
n_values <- c(300, 500, 700, 900, 1100, 1300, 1500)
results <- data.frame(n = integer(), method = character(), time = numeric())

# Loop over different data sizes
for (n in n_values) {
  scaled_data <- generate_data(n)
  
  # Measure time for K-Means
  tic()
  kmeans(scaled_data, centers = 3, nstart = 10)
  time_kmeans <- toc(log = TRUE)
  
  # Measure time for K-Medoids
  tic()
  pam(scaled_data, 3)
  time_kmedoids <- toc(log = TRUE)
  
  # Store results
  results <- rbind(results, data.frame(n = n, method = "K-Means", time = as.numeric(gsub(" sec elapsed", "", time_kmeans[[1]]))))
  results <- rbind(results, data.frame(n = n, method = "K-Medoids", time = as.numeric(gsub(" sec elapsed", "", time_kmedoids[[1]]))))
}

# Plot runtime comparison
ggplot(results, aes(x = n, y = time, color = method)) +
  geom_line(linewidth = 1) +  # Use linewidth instead of size
  geom_point(size = 2) +
  scale_x_log10() +  # Log scale for better visualization
  ggtitle("Runtime Comparison: K-Means vs. K-Medoids") +
  xlab("Number of Data Points (log scale)") +
  ylab("Time (seconds)") +
  theme_minimal()
```
